A generalization of the Bradley-Terry model for draws in chess

### 2019-05-20 inst/collusion_chess_draws.Rnw
minor wording

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}
\usepackage{units}
\usepackage{xcolor}

@@

\cite{davidson1970}

\cite{joe1990}

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won, thus reducing chess game to a
Bernoulli trial, almost surely.  This system became unworkable due to
the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

\cite{davidson1970} uses a threshhold approach to interpret a draw as
performance of the players the same to within a threshhold.  Using a
logistic expression for the difference in players' performance, he
effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  Here we
note that such a model automatically give stronger players a higher
draw potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear and likelihood maximization becomes
difficult.

@@

\begin{figure}[t]
\begin{center}
<<echo=FALSE,fig=TRUE>>=
dotchart(maxp(kka_3draws),pch=16)
@ 
\caption{Maximum likelihood estimate for \label{kka_3draws} equation \ref{KKA_personalised_drawmonster} equation} 
  \end{center}
\end{figure}

@@

The method of support allows one to account for proclivity to draw
while accounting for players' strengths and first-move advantage.  We
now consider one famous incident, following the World Chess
Championships 1963, in which Fischer famously and publicly claimed
that the Soviet players colluded.  The accusation was that Soviet
players agreed ahead of time to draw when playing one another, in
order to conserve energy for subsequent matches.

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis of Cura\c{c}ao.  Unlike
the present analysis, which uses a dataset restricted to game result
(win/loss/draw), \citet{moul2009} included number of moves in each
game\footnote{The reasoning was that hard-fought games that happened
  to end in a draw are more likely to last longer than collusive or
  pre-arranged draws.} in their analysis, and used a dataset of 7355
games played in the period 1940-1978 to calibrate the ability of the
players.  In contrast, the analysis presented uses a dataset
restricted to game outcome (win-loss-draw) in the competition itself.

@@

Here I analyse the original subject of the complaint, the Cura\c{c}ao
1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
of the Soviet players (Keres, Petrosian, and Geller) had agreed to
draw amongst themselves.  Observe that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher exact test, one-sided,
$p=0.0079$).  The likelihood function for this dataset would have

@@

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 36 terms such as
$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
$\left(P+\mathit{Ke}+W+C\right)^{4}$ on the denominator.  The
likelihood ratio for the hypothesis that collusive draws have the same
strength as regular draws isabout $e^{11.8}\simeq 10^5$: we conclude
that the high number of draws among the three accused players was
unlikely to be due to chance.

One plausible and charitable explanation for this result is that
Soviet players have a higher propensity to draw than western players.
We can investigate this by considering the interzonal tournament which
took place in Stockholm in 1962.

@@

For the results shown in table~\ref{interzonal}, we would have

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game has at least one
non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
and the likelihood function for table~\ref{interzonal} would be

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{units}

@@

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

@@

\cite{davidson1970} uses a threshhold approach to interpret a draw as
performance of the players the same to within a threshhold.  Using a
logistic expression for the difference in players' performance, he
effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  However,
\cite{joe1990} uses likelihood methods to show that this model does
not fit data well.  Here we note that such a model automatically give
stronger players a higher draw potential, and also note that it is
difficult to account for first-player advantage in this framework.

Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear and likelihood maximization becomes
difficult.

@@

\begin{figure}[t]
\begin{center}
<<echo=FALSE,fig=TRUE>>=
dotchart(maxp(kka_3draws),pch=16)
@ 
\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in
  equation  \ref{KKA_personalised_drawmonster}
} 
  \end{center}
\end{figure}

@@

The method of support allows one to estimate proclivity to draw while
accounting for players' strengths and first-move advantage.  We now
consider one famous incident, following the World Chess Championships
1963, in which Fischer famously and publicly claimed that the Soviet
players colluded.  The accusation was that Soviet players agreed ahead
of time to draw when playing one another, in order to conserve energy
for subsequent matches.

\citet{moul2009} consider this accusation from a statistical
perspective and present a detailed analysis.  Unlike the present work,
which uses a dataset restricted to game result (win/loss/draw),
\citet{moul2009} included number of moves in each game\footnote{The
  reasoning was that hard-fought games that happened to end in a draw
  are more likely to last longer than collusive or pre-arranged
  draws.} in their analysis, and used a dataset of 7355 games played
in the period 1940-1978 to calibrate the ability of the players.  In
contrast, the analysis presented uses a dataset restricted to game
outcome (win-loss-draw) in the competition itself.

@@

Here I analyse the original subject of the complaint, the Cura\c{c}ao
1962 Candidates (match results shown in Table~\ref{candidates}).
Fischer alleged that three of the Soviet players (Keres, Petrosian,
and Geller) had agreed to draw amongst themselves.  Observe that all
twelve of the games between the three accused players ended in a draw,
a significant result given 51 winning games and 54 draws (Fisher exact
test, one-sided, $p=0.0079$).

To investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet draw monster $C$, who wins when two colluding
players draw, and is inactive whenever a game has at least one
non-colluding player.  Then $H_N\colon C=D$ corresponds to no
collusion and the likelihood function would be

@@

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] on the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

One plausible explanation for this result is that Soviet players have
a higher propensity to draw than western players.  We can investigate
this by considering the interzonal tournament which took place in
Stockholm in 1962.

@@

The fith interzonal took place in Stockholm in 1962, in the form of a
23-player round-robin tournament.  For the results shown in
table~\ref{interzonal}, we would have a likelihood function

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of increased Soviet drawing proclivity, we
introduce a Soviet draw monster $D_s$, who wins when two Soviet
players draw, and is inactive if a game has at least one non-Soviet
player.  Then $H_N\colon D_s=D$ corresponds to no collusion and the
likelihood function for table~\ref{interzonal} would be
